A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-04 podc_paper/algorithm.tex
finished initial revision of linearizability proof
working on analysis section

Finally, when designing a total order for the operations propagated to the root, we choose
to put each block's enqueues before its dequeues.
Thus, if the root's blocks array contains blocks $B_1, \ldots, B_k$, the 
linearization ordering is 
\begin{equation}
L=E(B_1)\cdot D(B_1) \cdot E(B_2) \cdot D(B_2) \cdots E(B_k) \cdot D(B_k).
\end{equation}
\label{linearization}

@@

\Function{Object}{Dequeue()}{} 
    \State \hangbox{let \var{n} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
	    \fld{sum\sub{enq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}},\\
	    \fld{sum\sub{deq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1}\label{deqNew}
    \State \Call{Append}{\var{n}}
    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, 1}
    \State \Return{ \Call{FindResponse}{\var{b, i}}}
    \label{deqRest}
\EndFunction{Dequeue}

@@

\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    %  $i\geq 1$ and \var{root}.\fld{blocks}[\var{b}] is non-null and}
    %  \State \linecomment{contains at least \var{i} dequeues}
    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmpty}
        \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNull}
    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeE}
		\State \linecomment{compute enqueue's block using binary search}
		\State find min $\var{b}'\leq \var{b}$ with $\var{root}.\fld{blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{e}$
		\State \linecomment{find rank of enqueue within its block}
		\State $\var{i}' \assign\ \var{e} - \var{root}.\fld{blocks}[\var{b}'-1].\fld{sum\sub{enq}}$
        \State \Return \Call{GetEnqueue}{\var{root}, $\var{b}'$, $\var{i}'$}\label{findAnswer}
    \EndIf
\EndFunction{FindResponse}

@@

To compute the response of the $i$th \nf{Dequeue} in the $b$th block
of the root, \opemph{FindResponse}(\var{b, i}) determines at line \ref{checkEmpty} if the queue is empty.
If the queue is
not empty, Line \ref{computeE} computes the rank \var{e} of the
\nf{Enqueue} whose argument is the response to the
\nf{Dequeue}. 
A binary search is used on the \fld{sum\sub{enq}} fields of \var{root}.\fld{blocks} to find the index $b'$ of the block that contains 
the \var{e}th enqueue.
The index $b'$ is at most $b$ (since the enqueue is linearized before the dequeue).  To find the left end of the range for the binary search for $b'$, we can first do a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.
Then, \op{GetEnqueue} is used to trace down through the tree to find the required enqueue in a leaf.

>>>>

Finally, when designing a total order for the operations propagated to the root, we choose
to put each block's enqueues before its dequeues.
Thus, if the root's blocks array contains blocks $B_1, \ldots, B_k$, the 
linearization ordering is 
\begin{equation}
L=E(B_1)\cdot D(B_1) \cdot E(B_2) \cdot D(B_2) \cdots E(B_k) \cdot D(B_k).
\label{linearization}
\end{equation}

@@

\Function{Object}{Dequeue()}{} 
    \State \hangbox{let \var{n} be a new \typ{Block} with \fld{element} \assign\ \nl,\\
	    \fld{sum\sub{enq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}},\\
	    \fld{sum\sub{deq}} \assign\ \var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1}\label{deqNew}
    \State \Call{Append}{\var{n}}
    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, 1}\label{invokeIndexDequeue}
    \State \Return{ \Call{FindResponse}{\var{b, i}}}\label{deqRest}
\EndFunction{Dequeue}

@@

\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    %  $i\geq 1$ and \var{root}.\fld{blocks}[\var{b}] is non-null and}
    %  \State \linecomment{contains at least \var{i} dequeues}
    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}\label{FRNum}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmpty}
        \State \Return \nl \hfill \com\ queue is empty when dequeue occurs\label{returnNull}
    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeE}
		\State \linecomment{compute enqueue's block using binary search}
		\State find min $b_e\leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRb}
		\State \linecomment{find rank of enqueue within its block}
		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$\label{FRi}
        \State \Return \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswer}
    \EndIf
\EndFunction{FindResponse}

@@

To compute the response of the $i$th \nf{Dequeue} in the $b$th block
of the root, \opemph{FindResponse}(\var{b, i}) determines at line \ref{checkEmpty} if the queue is empty.
If the queue is
not empty, Line \ref{computeE} computes the rank \var{e} of the
\nf{Enqueue} whose argument is the response to the
\nf{Dequeue}. 
A binary search is used on the \fld{sum\sub{enq}} fields of \var{root}.\fld{blocks} to find the index $b_e$ of the block that contains 
the \var{e}th enqueue.
The index $b_e$ is at most $b$ (since the enqueue is linearized before the dequeue).  To find the left end of the range for the binary search for $b_e$, we can first do a doubling search \cite{BY76}, comparing \var{e} to the \fld{sum\sub{enq}} fields at indices $b-1, b-2, b-4, b-8, \ldots$.
Then, \op{GetEnqueue} is used to trace down through the tree to find the required enqueue in a leaf.

### 2023-01-04 podc_paper/analysis.tex
working on analysis section

\section{Analysis}
In this section, we analyze the number of \nf{CAS} invocations and the step complexity of our algorithm.

First, we prove some claims about the size and operations of a block. These lemmas will be used later for the correctness and analysis of \nf{GetEnqueue()}.

\begin{lemma}\label{blockSize}
Each block $B$ in each node contains at most one operation of each process.
If $c$ is the execution's maximum point contention, $B$ has at most $c$ direct~subblocks.
\end{lemma}
\begin{proof}
Suppose $B$ contains an operation of process $p$.
Let $op$ be the earliest operation by $p$ contained in $B$.
When $op$ terminates, $op$ is contained in $B$ by \Cref{lem::appendExactlyOnce}.
Thus, $B$ cannot contain any later operations by $p$, since $B$ is created before
those operations are invoked.

Let $t$ be the earliest termination of any operation contained in $B$.
By \Cref{lem::appendExactlyOnce}, $B$ is created before $t$, so all operations contained in $B$
are invoked before $t$.  Thus, they are all concurrently running at $t$, so there are at most
$c$ operations contained in $B$.
By definition, the direct subblocks of $B$ contain these $c$ operations, and each operation is contained
in exactly one of these subblocks, by Lemma \ref{lem::subblocksDistinct}.
By \Cref{blockNotEmpty}, each direct subblock of $B$ contains at least one operation,
so there are at most $c$ direct subblocks.
\end{proof}

\begin{proposition}
An \nf{Enqueue} or \nf{Dequeue} operation does at most $14\log p$ \nf{CAS} operations.
\end{proposition}
\begin{proof}
  In each level of the tree  \nf{Refresh} is invoked at most two times, and every \nf{Refresh} invokes at most seven \nf{CAS}es, one in Line \ref{cas} and two from each \nf{Advance} in Line \ref{helpAdvance} or \ref{advance}.
\end{proof}

\begin{lemma}[\nf{DoublingSearch} Analysis]\label{dsearchTime}
If the \nf{element} enqueued by $E_i(root,b)=E_e(root)$ is the response to some \nf{Dequeue} operation in \nf{root.blocks[$end$]}, then \nf{DoublingSearch($e$, $end$)} takes $O\big(\log ( \nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size})\big )$ steps.
\end{lemma}
\begin{proof}
First we show $end - b -1\leq 2 \times \nf{root.blocks[$b-1$].size}+\nf{root.blocks[$end$].size}$. There can be at most \nf{root.blocks[$b$].size}  \nf{Dequeue}s in \nf{root.blocks[$b+1\cdots end-1$]}; otherwise all elements enqueued by \nf{root.blocks[$b$]} would be dequeued before \nf{root.blocks[$end$]}. Furthermore, in the execution of queue operations in the linearization ordering, the size of the queue becomes \nf{root.blocks[$end$].size} after the operations of \nf{root.blocks[$end$]}. The final size of the queue after \nf{root.blocks[$1\cdots end$]} is \nf{root.blocks[$end$].size}. After an execution on a queue, the $size$ of the queue is greater than or equal to $\#enqueues -\#dequeues$ in the execution. We know the number of dequeues in \nf{root.blocks[$b+1\cdots end-1$]} is less than \nf{root.blocks[$b$].size}, therefore in \nf{root.blocks[$b+1\cdots end-1$]} there cannot be more than $\nf{root.blocks[$b$].size} + \nf{root.blocks[$end$].size}$ \nf{Enqueue}s. Overall there can be at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ operations in \nf{root.blocks[$b+1\cdots end-1$]} and since from Line \ref{addOP} we know that the \nf{num} field of every block in the tree is greater than 0, each block has at least one operation, so there are at most $2 \times\nf{root.blocks[$b$].size}+ \nf{root.blocks[$end$].size}$ blocks in between \nf{root.blocks[$b$]} and \nf{root.blocks[$end$]}. So, $end-b-1\leq 2 \times\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size}$.

Thus, the doubling search reaches \nf{start} such that the \nf{root.blocks[start].sum\sub{enq}} is less than $e$ in $O \big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$ steps. See Figure \ref{fig::doubling}. After Line \ref{dsearchEnd}, the binary search that finds $b$ also takes $O\big(\log(\nf{root.blocks[$b$].size}+\nf{root.blocks[$end$].size})\big)$. Next, \nf{i} is computed via the definition of \nf{sum\sub{enq}} in constant time (Line \ref{DSearchComputei}).
\end{proof}
\begin{figure}[hbt]  
  \center\includegraphics[width=6in]{pics/doubling.png}
  \caption{Distance relations between \nf{start}$,b,end$.}
  \label{fig::doubling}
\end{figure}

>>>>

\section{Analysis}
In this section, we analyze the number of steps and the number of \op{CAS} instructions performed by operations.
First, we prove some claims about the size and operations of a block.

\begin{proposition}
Each \nf{Enqueue} or \nf{Dequeue} operation performs $O(\log p)$ \nf{CAS} instructions.
\end{proposition}
\begin{proof}
An operation invokes \nf{Refresh} at most twice at each of the $\ceil{\log_2 p}$ levels of the tree.
A \nf{Refresh} does at most 7 \nf{CAS} steps: one in line \ref{cas} and two from each \nf{Advance} in line \ref{helpAdvance} or~\ref{advance}.
\end{proof}

\begin{lemma}\label{dsearchTime}
The search that \op{FindResponse}$(b,i)$ does at line \ref{FRb} to find the index $b_e$ of the block in the root containing the $e$th enqueue takes $O(\log (\var{root}.\fld{blocks}[b_e].\size + \var{root}.\fld{blocks}[b-1].\size))$ steps.
\end{lemma}
\begin{proof}
Let the blocks in the root be $B_1, \ldots, B_\ell$.
The doubling search used to compute $b_e$ takes $O(\log (b-b_e))$ steps,
so we prove $b - b_e \leq 2 \cdot B_{b_e}.\size + B_{b-1}.\size + 1$.
If $b \leq b_e+1$, then this is trivial, so assume for the rest of the proof that $b>b_e+1$.

As shown in \Cref{linearCorrect}, the dequeue that calls \op{FindResponse} is in $B_b$ and is supposed to return an enqueue in $B_{b_e}$.
Thus, there can be at most $B_{b_e}.\size$ dequeues in 
$D(B_{b_e+1}) \cdots D(B_{b-1})$; otherwise in the sequential execution $L$,
all elements enqueued before the end of
$E(B_{b_e})$ would be dequeued before $D(B_b)$. 
Furthermore, by \Cref{sizeCorrectness}, the size of the queue  after the prefix of $L$ corresponding to 
$B_1,\ldots,B_{b-1}$  is 
$B_{b-1}.\size \geq B_{b_e}.size + |E(B_{b_e+1})\cdots E(B_{b-1})| - |D(B_{b_e+1})\cdots D(B_{b-1})|$.
Thus, $|E(B_{b_e+1})\cdots E(B_{b-1})| \leq B_{b-1}.\size + |D(B_{b_e+1})\cdots D(B_{b-1})| \leq B_{b-1}.\size + B_{b_e}.\size$.
So, the total number of operations in $B_{b_e+1}, \ldots, B_{b-1}$ is at most
$B_{b-1}.\size + 2\cdot B_{b_e}.\size$.
Each of these $b-1-b_e$ blocks contains at least one operation, by \Cref{blockNotEmpty}.
So, $b-1-b_e \leq B_{b-1}.\size + 2\cdot B_{b_e}.\size$
\end{proof}

The following lemma is useful in bounding the time to \op{GetEnqueue}
\begin{lemma}\label{blockSize}
Each block $B$ in each node contains at most one operation of each process.
If $c$ is the execution's maximum point contention, $B$ has at most $c$ direct~subblocks.
\end{lemma}
\begin{proof}
Suppose $B$ contains an operation of process $p$.
Let $op$ be the earliest operation by $p$ contained in $B$.
When $op$ terminates, $op$ is contained in $B$ by \Cref{lem::appendExactlyOnce}.
Thus, $B$ cannot contain any later operations by $p$, since $B$ is created before
those operations are invoked.

Let $t$ be the earliest termination of any operation contained in $B$.
By \Cref{lem::appendExactlyOnce}, $B$ is created before $t$, so all operations contained in $B$
are invoked before $t$.  Thus, all are  running concurrently at $t$, so $B$ contains at most $c$ operations.
By definition, the direct subblocks of $B$ contain these $c$ operations, and each operation is contained
in exactly one of these subblocks, by Lemma \ref{lem::subblocksDistinct}.
By \Cref{blockNotEmpty}, each direct subblock of $B$ contains at least one operation,
$B$ has at most $c$ direct subblocks.
\end{proof}

--------

### 2023-01-04 podc_paper/correctness.tex
working on analysis section

\subsection{Correctness of GetEnqueue}

@@

\subsection{Correctness of IndexDequeue}

@@

\begin{lemma}\label{lem::indexDequeue}
If $v.\fld{blocks}[b]$ has been propagated to the root and $1\leq i\leq |D(v.\fld{blocks}[b])|$, 
 then \op{IndexDequeue}($v, b, i$) returns $\langle b',i'\rangle$ such that the \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is the $(i')$th dequeue of $D(\var{root}.\fld{blocks}[b'])$.
\end{lemma}
\begin{proof}
We prove the claim by induction on the depth of node $v$. The base case where $v$ is the root is trivial (see Line \ref{indexBaseCase}).
Assuming the claim holds for $v$'s parent, we prove it for $v$.
Let $B=v.\fld{blocks}[b]$ and $B'$ be the superblock of $B$.
\op{IndexDequeue}($v, b, i$) first computes the index $sup$ of $B'$ in $v.\fld{parent}$.
By \Cref{superRelation}, this index is either $B.super$ or $B.super+1$.
The correct index is determined by testing on line \ref{supertest} whether $B$ is not a subblock of $v.\fld{parent.blocks}[B.\fld{super}]$.

@@

\begin{lemma}\label{linearCorrect}
Each terminating dequeue returns the response it would in the sequential execution $L$.
\end{lemma}
\begin{proof}
If a dequeue $D$ terminates, it is contained in some block in the root, by \Cref{lem::appendExactlyOnce}.
By \Cref{lem::indexDequeue}, $D$'s call to \op{IndexDequeue} on line \ref{invokeIndexDequeue}
returns a pair $\langle b,i\rangle$ such that $D$ is the $i$th dequeue in the block 
$B=\var{root}.\fld{blocks}[b]$.
$D$ then calls \op{FindResponse}($b,i$) on line \ref{deqRest}.
By \Cref{sizeCorrectness}, the queue contains $\var{root}.\fld{blocks}[b-1].\fld{size}$ elements
after the operations in $\var{root}.\fld{blocks}[1..b-1]$ are performed sequentially 
in the order given by $L$.
By \Cref{lem::sum}, the value of \var{num\sub{enq}} computed on line \ref{FRNum}
is the number of enqueues in $B$.
Since the enqueues in block $B$ precede the dequeues,
the queue is empty when the $i$th dequeue of $B$ occurs if 
$\var{root}.\fld{blocks}[b-1].\fld{size} + \var{num\sub{enq}} < i$.
So $D$ returns \nl\ on line \ref{returnNull} if and only if it would do so in the sequential
execution $L$.
Otherwise, the size of the queue after doing the operations in $\var{root}.\fld{blocks}[0..b-1]$
in the sequential execution $L$ is $\var{root}.\fld{blocks}[b-1].\fld{sum\sub{enq}}$ minus
the number of non-\nl\ dequeues in that prefix of $L$.
Hence, line \ref{computeE} sets $e$ to the rank of $D$ among all the non-\nl\ dequeues in $L$.
Thus, in the sequential execution~$L$, $D$ returns the value enqueued by the $e$th enqueue in $L$.
By \Cref{lem::sum}, this enqueue is the ($i'$)the enqueue 
in $E(\var{root}.\fld{blocks}[b'])$, where
$b'$ and $i'$ are the values $D$ computes on line \ref{FRb} and \ref{FRi}.
By \Cref{lem::get}, the call to \op{GetEnqueue} returns the argument of the required enqueue.
\end{proof}

Combining \Cref{linearSat} and \Cref{linearCorrect} gives us our main result.

>>>>

\subsection{Correctness of Helping Routines}

We first show that \op{getEnqueue} works correctly.

@@

@@

\begin{lemma}\label{lem::indexDequeue}
If $v.\fld{blocks}[b]$ has been propagated to the root and $1\leq i\leq |D(v.\fld{blocks}[b])|$, 
 then \op{IndexDequeue}($v, b, i$) returns $\langle b',i' \rangle$ such that the \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is the $(i')$th dequeue of $D(\var{root}.\fld{blocks}[b'])$.
\end{lemma}
\begin{proof}
We prove the claim by induction on the depth of node $v$. The base case where $v$ is the root is trivial (see Line \ref{indexBaseCase}).
Assuming the claim holds for $v$'s parent, we prove it for $v$.
Let $B=v.\fld{blocks}[b]$ and $B'$ be the superblock of $B$.
\op{IndexDequeue}($v, b, i$) first computes the index $sup$ of $B'$ in $v.\fld{parent}$.
By \Cref{superRelation}, this index is either $B.super$ or $B.super+1$.
The correct index is determined by testing on line \ref{supertest} whether $B$ is not a subblock of $v.\fld{parent.blocks}[B.\fld{super}]$.

@@

Next, we show operations return the same response as they would in the sequential execution $L$.

\begin{lemma}\label{linearCorrect}
Each terminating dequeue returns the response it would in the sequential execution $L$.
\end{lemma}
\begin{proof}
If a dequeue $D$ terminates, it is contained in some block in the root, by \Cref{lem::appendExactlyOnce}.
By \Cref{lem::indexDequeue}, $D$'s call to \op{IndexDequeue} on line \ref{invokeIndexDequeue}
returns a pair $\langle b,i\rangle$ such that $D$ is the $i$th dequeue in the block 
$B=\var{root}.\fld{blocks}[b]$.
$D$ then calls \op{FindResponse}($b,i$) on line \ref{deqRest}.
By \Cref{sizeCorrectness}, the queue contains $\var{root}.\fld{blocks}[b-1].\fld{size}$ elements
after the operations in $\var{root}.\fld{blocks}[1..b-1]$ are performed sequentially 
in the order given by $L$.
By \Cref{lem::sum}, the value of \var{num\sub{enq}} computed on line \ref{FRNum}
is the number of enqueues in $B$.
Since the enqueues in block $B$ precede the dequeues,
the queue is empty when the $i$th dequeue of $B$ occurs if 
$\var{root}.\fld{blocks}[b-1].\fld{size} + \var{num\sub{enq}} < i$.
So $D$ returns \nl\ on line \ref{returnNull} if and only if it would do so in the sequential
execution $L$.
Otherwise, the size of the queue after doing the operations in $\var{root}.\fld{blocks}[0..b-1]$
in the sequential execution $L$ is $\var{root}.\fld{blocks}[b-1].\fld{sum\sub{enq}}$ minus
the number of non-\nl\ dequeues in that prefix of $L$.
Hence, line \ref{computeE} sets $e$ to the rank of $D$ among all the non-\nl\ dequeues in $L$.
Thus, in the sequential execution~$L$, $D$ returns the value enqueued by the $e$th enqueue in $L$.
By \Cref{lem::sum}, this enqueue is the $i_e$th enqueue 
in $E(\var{root}.\fld{blocks}[b_e])$, where
$b_e$ and $i_e$ are the values $D$ computes on line \ref{FRb} and \ref{FRi}.
By \Cref{lem::get}, the call to \op{GetEnqueue} returns the argument of the required enqueue.
\end{proof}

Combining \Cref{linearSat} and \Cref{linearCorrect} provides our main result.
